### 2020-07-03 manuscript/manuscript.rmd
finished first full pass at main text

There are other approaches to addressing heterogeneity in equilibrium frequencies. For example, Feuda *et al.* [@Feuda:2017ew] emplyed recoding methods that recoded the full set of twenty amino acids into six groups of amino acids. These groups have more frequent evolutionary changes within them than between them, based on empirical observations in large protein datasets [@Susko:2007ds]. This approach could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew]. Our own analyses (Supplementary text, Supplementary Fig. XX-XX) indicate, however that the impact of data recoding is is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that it is premature to accept Porifera-sister and reject Ctenophora-sister using data-recoding methods and recoding can be a problematic method for addressing compositional variation.

## The current state of understanding and future direction
Resolving the placement of the root in the animal phylogeny has proven particularly difficult, even with large genome-scale phylogenomic data matrices. By synthesizing past phylogenomic studies and performing new analyses to characterize difference between studies, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with sampling restricted to the most closely related outgroups. Four factors contribute significantly to the previous incongruent results based on our analyses. First, the overlap of gene sampling and matrix is extremely low across different studies, leading to the difficult of directly compare the results across different studies. Second, the distribution of phylogenetic signal are largely varied between site-homogeneous and site-heterogeneous models. More importantly, site-heterogeneous models are also more sensitive to outgroup choice than site-homogeneous models, leading to different results even with the same data matrix with different outgroup choice. Third, we found that increase of number of substitutional categories coupled with a more simplistic model can change the root position of animal phylogeny. Last, we demonstrated that the support of Porifera-sister by data-recoding methods is potentially resulted by data reduction, not accommodation of compositional heterogeneity across species. Thus, the recovery of Porifera-sister needs more constraints than Ctenophora-sister.

Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised. Overcoming the challenges to inferring the earliest diverging animal lineage is particular difficult and here we discuss the several aspects that will help to improve our understanding of animal evolution. First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available). Second, accurate reconstruction of orthologs and a consensus analytical strategy should be formalized with investigators to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated]. Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem. More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work. Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian]. We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.

>>>>

Other approaches have been taken to addressing heterogeneity in equilibrium frequencies. For example, Feuda *et al.* [@Feuda:2017ew] employed recoding methods that recoded the full set of twenty amino acids into six groups of amino acids. These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds]. Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew]. Our own analyses (Supplementary text, Supplementary Fig. XX-XX) indicate, however that the impact of data recoding is is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that recoding can be a problematic method for addressing heterogeneity.

## The current state of understanding and future direction

Resolving the placement of the root in the animal phylogeny has proven particularly difficult. By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some analyses. This demonstrates and important interaction between model selection and taxon sampling. Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency. Site homogenous models have a single category that is applied to all sites. C models, as implemented in IQ-TREE, have 10-16 categories. CAT models can estimate the number of categories, in which case they use more than 500 for the matrices assessed here, or set to a specific number with nCAT. When outgroup sampling is fixed to a restricted subset (a requirement for any analysis to support Porifera-sister), root position is dependent on the number of categories. As the number of categories is increased, at some point support for Ctenophora-sister diminishes and support for Porifera-sister increases (Fig. 4). As part of this process, the number of genes with phylogenetic signal for each hypothesis also changes.

This dependence alone doesn't resolve the question of where the root falls in the animal tree of life, that depends on how we interpret this dependence. Focusing just on heterogeneous analyses of matrices with restricted outgroup sampling, should we believe the results with fewer equilibrium frequency categories or the results with many? XX

Overcoming the challenges that face this difficult and important question goes beyond models. In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic perspective. In particular, there are very few chromosome-level genome assembles outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. This is is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and greatly reduce the possibility of contamination in any of the gene sequence datasets. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
